---
title: "Template for summarized heatmaps"
output: html_notebook
---
```{r echo=FALSE, messages=FALSE}
library(elsasserlib)
library(ggplot2)
library(reshape2)
library(ggpubr)
# Ongoing Issue with mcols <-
library(rtracklayer)
```

This is a template markdown file that creates a summary heatmap (like the ones
we make for ChromHMM17)

## Calculating the aggregated values

This is the most computationally costful part. I recommend to put results in a
table after running this if you want to play with them. Additionally, this notebook
contains a parameter `cache=TRUE` for the costful chunks of code, which means
if you don't touch those chunks after running, it will not try to run them
again.

If you want, first set your working directory to the one where the data is:

```{r}
expdir <- 'path-to-your-experiment'

```

Now this will plot and label every bigwig file on a certain directory:

```{r}

run.dir <- './bw'
bwpath <- paste(expdir, run.dir, sep='/')

# You can actually do this with any other type of bed file. The important part
# is that the name field contains categories that can be aggregated. For instance
# you could also run genes_hi / genes_med and so on
bed <- 'path-to-chromhmmfile-or-any-other-bed'

# This lists all the files in the path you put there
bwfiles <- list.files(bwpath)

# And outputs them so you can see if there is any problem with it.
bwfiles

```

Now we calculate the aggregated values across all these files, using
true mean:

```{r cache=TRUE}
values <- bw_bed(paste(bwpath, bwfiles, sep='/'),
                 bed,
                 aggregate.by='true_mean')
```

We get a `data.frame` object where each column is a file and each row a ChromHMM
value:
```{r}
head(values)
```

Reorder the values:

```{r}
library(stringr)

# This is a trick for naturally sort the values when they have numbers but they
# don't have padded zeros to the left. Basically applies to ChromHMM values
order <- str_sort(values$name, numeric=TRUE)

rownames(values) <- values$name
write.csv(values[order, bwfiles], 'your_bedfile_truemeans.csv')
```

## Plotting the tables

This uses `pheatmap` and `RColorBrewer` to plot the values. I will add this
functionality to the package, but as of now, here is code that will do. You
may need to adapt `fig.width` and `fig.height` on the header, as pheatmap does 
not seem to be aware of the limits of the page.

```{r fig.width=12, fig.height=7}
library(pheatmap)
library(RColorBrewer)

# This computes limits and breaks for the color scale to look always the same,
# regardless of only positive values or -max, 0, +max.
compute.lims <- function(mat) {
  maxmat <- mat
  maxmat[is.na(maxmat)] <- -Inf
  maxval <- max(maxmat)

  minmat <- mat
  minmat[is.na(minmat)] <- Inf
  minval <- min(minmat)

  nsteps <- 21.0

  breaklim <- ceiling(abs(max(abs(c(minval, maxval)))))

  # Compute a reasonable amount of steps?
  stepsize <- 2*breaklim / nsteps

  breakslist <- seq(-breaklim, +breaklim, by=stepsize)
  breakslist
}

# Just a wrapper that sets some defaults on pheatmap
summary_heatmap <- function(values, title, size=35) {
  bcolor <- "white"

  breakslist <- compute.lims(values)
  palette <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(length(breakslist))
 
  cellsize <- size
  
  pheatmap(values,
           main=title,
           cellwidth=cellsize,
           cluster_rows=F,
           cluster_cols=F,
           cellheight=cellsize,
           border_color=bcolor,
           breaks=breakslist,
           color=palette,
           display_numbers=TRUE)
}


values.from.table <- read.csv('./chromhmm17_truemean_runs.csv', header=T, row.names = 1)
summary_heatmap(t(values.from.table), "Log2 True mean per category")
```

## Checking other aggregating functions

We can check variability for median or mean (mean-of-means) values.

```{r cache=TRUE}
median.values <- bw_bed(paste(bwpath, bwfiles, sep='/'),
                 bed,
                 aggregate.by='median')

```

```{r}
rownames(median.values) <- median.values$name

write.csv(median.values[order, bwfiles], 'your_bedfile_median_runs.csv')

median.values[order, bwfiles]
```

You can plot also the median values:

```{r fig.width=12, fig.height=7}
summary_heatmap(t(median.values[order, bwfiles]), "Median mean per category")
```
