---
title: "Elsasserlib BigWig functionality"
author: "Carmen Navarro"
date: "Nov 12 2020"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{plot-bigwig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{elsasserlib}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Summary

This is a brief documentation on how to use the `elsasserlib` package to visualize
the content of `bigWig` files. There are several ways you can visualize this kind
of data, which can be grouped into two main categories:

- **Genome-wide analysis**. Genome is partitioned on equally-sized bins and
their value distributions are plotted. Useful to get a general idea of the
signal distribution without looking at specific places.
- **Across a set of *loci***. One can be interested in looking at a specific set
of *loci*. This can be either summarized categories, profile plots or heatmaps.

# About the built-in data

This package comes with a set of small files to show the functionality. These
have been built from a published data from our lab and correspond to a 
H3.3 and H3K9me3 ChIP + input (`GSE149080`), subset across a
500kbp genomic region: `chr15-102600000-103100000`, which overlaps with the *HOXC*
gene cluster. A ChromHMM annotation has also been subsampled to overlap with
such region.

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(scales)
library(rtracklayer)
library(elsasserlib)

h33_chip <- system.file("extdata", "sample_H33_ChIP.bw", package = "elsasserlib")
h3k9me3_chip <- system.file("extdata", "sample_H3K9me3_ChIP.bw", package = "elsasserlib")
input_chip <- system.file("extdata", "sample_Input.bw", package = "elsasserlib")
genes <- system.file("extdata", "sample_genes_mm9.bed", package = "elsasserlib")
chromhmm <- system.file("extdata", "sample_chromhmm.bed", package = "elsasserlib")
```

All these values are paths to bigWig and BED files.

# Genome-wide plots

Bin-based genome-wide analysis is currently supported in two formats: 

- `plot_bw_bins_violin`. Makes a violin plot of one or more bigWig files.
- `plot_bw_bins_scatter`. Makes a scatter plot of two different bigWig files
against each other. 

Both plot types allow for background bigWig signal to be provided (i.e. input).

## Violin plots

For a set of bigWig files, it is possible to visualize the genome-wide signal
distribution. For example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_violin(h33_chip, bin_size = 50000)
```

Note that `bin_size` parameter influences the time any of these functions take
to run. Reasonable resolution for this is 5000 or 10000.

It is possible to plot several bigWig files at a time:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_violin(c(h33_chip, input_chip), bin_size = 50000)
```

Additionally, one can highlight bins that overlap with a certain set of interest
loci. This is done via the `highlight` parameter:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_violin(
  c(h33_chip, input_chip),
  bin_size = 50000,
  highlight = genes,
  highlight_color = c("red", "red") # It is possible to provide different colors
)
```

It is also possible to provide a list of bigWig files to be used as background
to normalize the bin values to. For instance, in the previous case, one could
want to use the input values to normalize the H3.3 ChIP data:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
# bg_bwfiles parameter is set now to input_chip
plot_bw_bins_violin(
  h33_chip, 
  bg_bwfiles = input_chip,
  bin_size = 50000,
  highlight = genes,
  highlight_color = c("red")
)
```

If `norm_func` is provided, it is possible to log-transform these values as
well:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
# bg_bwfiles parameter is set now to input_chip
plot_bw_bins_violin(
  h33_chip, 
  bg_bwfiles = input_chip,
  bin_size = 50000,
  highlight = genes,
  highlight_color = c("#000088"), # Colors can be also HTML codes
  norm_func = log2  # Note this is a function, not a string
)
```

## Genome-wide scatterplots

Is it possible to compare two signals against each other on a scatter plot. To
do so, you can use `plot_bw_bins_scatter`. This function works very similarly to
`plot_bw_bins_violin` but instead of accepting an array of bigWig files, it
accepts separate `x` and `y` parameters and the corresponding optional 
`bg_x` and `bg_y`, `norm_func_x`, `norm_func_y`. You can see more details in
the function documentation.

One example would be looking at H3.3 versus H3K9me3 in this sample data:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_scatter(
  x = h33_chip,
  y = h3k9me3_chip,
  bin_size = 50000
)
```

This is useful to see relationships between marks. In this case, as it usually
happens, bins where H3.3 signal is high, H3K9me3 signal is low and viceversa.

The same way as with `plot_bw_bins_violin`, it is possible to specify a set of
loci to highlight within the bins, and normalize to given signals:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=7}
plot_bw_bins_scatter(
  x = h33_chip,
  bg_x = input_chip,
  y = h3k9me3_chip,
  bg_y = input_chip,
  bin_size = 50000,
  highlight = genes,
  highlight_colors = "red",
  highlight_label = "genes", # It is also possible to label the groups
  norm_func_x = log2 # Each norm func is provided separately
)
```

# Locus-specific plots

It is also possible to look at the ChIP data in a more specific way. For this,
`elsasserlib` provides the following functions:

- `plot_bw_bed_summary_heatmap`: Summarizes the values across a BED file, using
name field as category. 
- `plot_bw_profile`: Creates a profile plot across a set of loci.
- `plot_bw_heatmap`: Creates a heatmap across a set of loci.

## Summary heatmaps

These plots are meant to provide a genome-wide overview of the signal across
different biologically-relevant categories. They get a list of bigWig files
and a BED file as input. Each bigWig file will represent a row in the heatmap,
and each cell will be an aggregate value of the corresponding bigWig file in
a given category in the BED file, for example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=3}
# ChromHMM is a genome-wide annotation according to epigenetics marks. Each
# locus is tagged by a category. And the amount of categories must be limited.
# In this case, it is fifteen. 
chrom_values <- import(chromhmm, format = "BED")
head(chrom_values)

plot_bw_bed_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  bedfile = chromhmm,
)
```

Since this example is reduced to a set of HOXC genes, you can see how H3.3
signal is high and H3K9me3 is generally low.

It is possible as in the previous types of plots to provide bigWig files as
normalization / input tracks and use them in the same way as before:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=3}
plot_bw_bed_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  bg_bwfiles = c(input_chip, input_chip),
  bedfile = chromhmm
)
```

### How to save summary heatmap plots

These plots are made under the hood with the nice `pheatmap` library, which is
a dependency of the package. Since these do not output `ggplot` objects, 
`ggsave` will not work. To this end, there is a parameter that can be passed
to the function: `plot_bw_bed_summary_heatmap(..., file_out = "output.pdf")`.

### How the values are aggregated

Since this function is aggregating the values, it is possible to specify the 
way the per-locus values are aggregated. Default is **true mean**. What it means
is that all the coverage values are binned together and the mean is performed
as if each category is one big locus. Another possibilities are:
**mean of means** (`aggregate_by = "mean"`), which will calculate mean coverage
per locus and then the aggregate value will be the mean of those mean values,
and **median of means** (`aggregate_by = "median"`). Median per-locus coverage
values tend to be noisy, so a median of means could be useful in some cases.

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=3}
plot_bw_bed_summary_heatmap(
  c(h33_chip, h3k9me3_chip),
  bg_bwfiles = c(input_chip, input_chip),
  bedfile = chromhmm,
  aggregate_by = "median"   # Note this is a string value, not a function
)
```

## Profile plots

Another way to look at how the ChIP data looks is to summarize a profile across
a set of *loci*. In order to do that you can use the `plot_bw_profile` function:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input")
)
```

The default mode is called `stretch`, where start (3' end) and end (5' end) of
*loci* are aligned and the length of the *locus* is stretched to a given 
pseudolength (`middle` parameter). If this parameter is not given, the pseudolength
will be assigned as the median value of the length distribution of *loci*, 
in an attempt to match a realistic length. 

It is possible to align only 3' ends, 5' ends or center of loci, similarly to
what formatting text alignment would do:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input"),
  mode = "start"
)
```

It is also possible to specify the upstream and downstream lengths:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input"),
  mode = "center",
  upstream = 3000,
  downstream = 3000
)
```

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes,
  labels = c("H3.3", "H3K9me3", "Input"),
  mode = "end",
  upstream = 7000,
  downstream = 3000,
  show_error = TRUE # Turn error estimates on
)

```

## Heatmaps

Finally, it is possible to look at the profile across a set of *loci* in a 
per-*locus* way by looking at heatmaps. 

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes
)
```

This gene annotation includes some variants of the same genes that overlap a lot,
so some of the rows look very similar or even like duplicates.

Most of the parameters are shared between `plot_bw_profile` and `plot_bw_heatmap`
making it possible to arrange them on top of each other later on, so it is
possible to align them in different ways:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  mode = "start",
  upstream = 2500,
  downstream = 2500
)
```

It is possible to specify `bin_size` to increase or decrease resolution (this
can also be done on profile plots):

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  mode = "start",
  upstream = 2500,
  downstream = 2500,
  bin_size = 20
)
```

At this point, heatmap values are shown in order from maximum mean value across
all the beans to lowest.

By default, colorscale is set to percentiles 0.01 and 0.99 of whole matrix 
range. You can set these values to fixed numbers. This is useful mostly to 
plot together several heatmaps with the same color scale. For example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  mode = "start",
  upstream = 2500,
  downstream = 2500,
  bin_size = 100,
  zmin = 2,
  zmax = 10
)
```

It is also possible to change the color of the heatmap using any `RColorBrewer`
available palette:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
plot_bw_heatmap(
  h33_chip,
  bedfile = genes,
  bin_size = 100,
  cmap = "RdYlBu"
)
```

There is also a `max_rows_allowed` parameter that defaults to 10000. This 
allows to plot large matrices by binning vertically the heatmap and plotting the
underlying mean associated, provided that a screen or a reasonable screen will
not be able to display all the data points anyway. If the number of rows is 
larger than `max_rows_allowed`, rows will be binned in a way that the resulting
plotted matrix will have `max_rows_allowed` height and a warning will be shown.
Since the matrix is not subsampled only, but the underlying values averaged,
this may result in a smoothened effect in the image. This is a similar approach
to how image resizing works.

# Comments on runtime and resolution options

All these functions work on genome-wide data, and often you will want to run
these on more than one bigWig file at a time. It is possible to run all of this
in a regular laptop, however if resolution is too high, waiting times will 
raise to minutes and even hours, depending on the amount of files and the 
given resolution.

In a intel i7 laptop, bins analyses for a single bigWig file in resolution
around 10000bp tend to take a few seconds. 5000bp is still reasonable interactive
time. For plotting under 5000 bp resolution you will need to wait quite some
time and I would recommend running these in a script outsinde R environment.

Locus-based analyses runtime tends to be smaller since the amount of values to
be calculated is smaller than genome-wide bins. An exception to this are 
ChromHMM plots, since these *are* also genome-wide bins, if only of different
lengths and labeled with categories. Keep this in mind when plotting a large
set of bigWig files. It will take some time as well.

# Customizing plots

All plotting functions (except for `plot_bw_bed_summary_heatmap`) return a 
`ggplot` object. This means that essentially you can overlay any other data
you want or further customize the results. For example:

```{r, warning = FALSE, message = FALSE, fig.width=9, fig.height=5}
profile_plot <- plot_bw_profile(
  c(h33_chip, h3k9me3_chip, input_chip),
  bedfile = genes
)

profile_plot + 
  ggtitle("New title for the plot") + # Add a custom title
  geom_hline(yintercept = 1, linetype="dashed") # Add a horizontal line on y=1
```

# Advanced use of elsasserlib

If none of these functions provide you with what you want, this package also
offers a set of functions that just return the underlying values of these plots.
This way, you can perform more advanced operations if you want. When possible,
these functions return a `GRanges` object. When the data is summarized in some
way, they return data frames.

Functions you can use to calculate values:

- `bw_bins`. Returns a `GRanges` object with the summarized values for each
gene in a given genome. If a list of bigWig files is provided, it returns a 
single `GRanges` object with one metadata column per bigWig file. By default,
these columns will be named after the bigWig file names. Otherwise you can
provide an array of custom names for it.
- `bw_bed`. It works the same as `bw_bins` but instead of partitioning the
genome in bins, it takes a BED file as a parameter and returns a `GRanges`
with score values corresponding to each locus. It is possible to summarize
this with extra parameters and obtain the values that are plotted in 
`bw_bed_summary_heatmap` function.
- `bw_profile`. Returns a data frame with summarized values for a profile plot,
including error estimates.
- `bw_heatmap`. Returns a matrix with the underlying values of the heatmap 
plot obtained by `plot_bw_heatmap` function.
