---
title: "Using plot themes"
author: "Carmen Navarro"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{plot-utils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{elsasserlib}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(elsasserlib)
library(ggplot2)
library(scales)
```

## Summary

This is a brief documentation on how to use the plot themes included in the
package. The idea is that this can help us keep a consistent *look and feel* 
when getting plots ready for publication. The most relevant need that triggered
this was that we need pretty huge font sizes for the print panel version of the
plots.

## Example

So as an example let's take the usual `iris` dataset and plot it normally using
ggplot:

```{r fig.width=7, fig.height=7} 
normal.plot <- ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length)) + 
  geom_point() + 
  ggtitle("Your usual default ggplot scatterplot")

normal.plot
```

You can modify the default behavior in your plot by adding one of the available
themes there are.

```{r fig.width=7, fig.height=7} 
normal.plot + 
  ggtitle("Different looks for same plot") +
  theme_elsasserlab_screen()

```

Just some tweaking to the looks of the plot, because I like more minimalistic
plots. If you want to use the `print` version, you just need to call `theme_elsasserlab_print()`
after:

```{r fig.width=7, fig.height=7} 
normal.plot + 
  ggtitle("Large fonts") +
  theme_elsasserlab_print()

```

## Customizing base themes

You can use the parameter `base_size` to tune the size of things in both screen
and print plots.

```{r fig.width=7, fig.height=7}

# Same plot with slightly less huge fonts
normal.plot + 
  ggtitle("Not-so-large fonts") +
  theme_elsasserlab_print(base_size=18)

```

In general, it's pretty simple and works as any `ggplot2` operation. You can
still override specifics if you need. Say that you want to rotate axis labels
because you have long text there.


```{r fig.width=7, fig.height=7}
normal.plot + 
  ggtitle("Back to screen theme + rotated axis labels") +
  theme_elsasserlab_screen() +
  theme(axis.text.x = element_text(angle=90, hjust=1))

```

## Custom palette

This is something I did for myself because I did not quite like the colorsets
for qualitative datasets I found. So I manually picked a few that I think play
nice together and can be fairly distinguishable from each other.

`palette_categorical(n)` will give you a list of up to 12 colors for categorical
data. If you ask for more than n though, it will provide something very similar
to default ggplot hue scale.

```{r fig.width=7, fig.height=7}
categorical.plot <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) +
  geom_point() +
  theme_elsasserlab_screen() +
  scale_color_manual(values=palette_categorical(3))

categorical.plot
```

These are the colors included:
```{r fig.width=7, fig.height=7}
show_col(palette_categorical(12))
```

